---
title: "One-way ANOVA Exercise"
author: "Put Your Name Here"
date: "2024-02-01"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

## Research Topic

Memory of a traumatic event becomes consolidated within hours. Intrusive
memories can then flash back repeatedly into the mind's eye and cause
distress. We investigated whether reconsolidation - the process during
which memories become malleable when recalled - can be blocked using a
cognitive task and whether such an approach can reduce these unbidden
intrusions. We predicted that reconsolidation of a reactivated visual
memory of experimental trauma could be disrupted by engaging in a
visuospatial task that would compete for visual working memory
resources. We showed that intrusive memories were virtually abolished by
playing the computer game Tetris following a memory-reactivation task 24
hr after initial exposure to experimental trauma. Furthermore, both
memory reactivation and playing Tetris were required to reduce
subsequent intrusions (Experiment 2), consistent with
reconsolidation-update mechanisms. A simple, non-invasive cognitive-task
procedure administered after emotional memory has already consolidated
(i.e., \> 24 hours after exposure to experimental trauma) may prevent
the recurrence of intrusive memories of those emotional events.

**Research Question**: Is there of effect of cognitive games on the
experience of intrusive memories?

**Paper data is based on**:
<https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4526368/>

***Our task***: Run a One-Way ANOVA to investigate this RQ.

```{r setup, echo=TRUE, message = FALSE}


library(patchwork)
library(car)
library(broom)
library(afex)
library(tidyverse)
library(pwr) 
library(lsr)
library(emmeans)
library(performance)


```

## Create your folder and set your working directory

Create a folder called `anova` in your folders for this class. Set it as
your working directory.

**Click Session -\> Set Working Directory -\> Choose Directory**

## Call in our Data

Download the **`James Holmes_Expt 2_DATA.csv`** dataset from Canvas, and
then call it into R, assigning it to the variable **`dat1`**. Use the
**`head()`** and **`summary()`** functions to inspect the data.

```{r}

dat1 <- read.csv("James Holmes_Expt 2_DATA.csv")

#use head()

#use summary()

```

## Run the following code to clean it

You do not need to worry about data wrangling this week. This code will
do it for use.

```{r}

dat2 <- dat1 %>% 
  mutate(subject = row_number()) %>% 
  select(subject, Condition, Days_One_to_Seven_Image_Based_Intrusions_in_Intrusion_Diary) %>% 
  rename("intrusions" = "Days_One_to_Seven_Image_Based_Intrusions_in_Intrusion_Diary")

```

## Identify your IV and your DV

Before we explore our data, we need to tell `R` which what are our
independent variable and dependent variables. Doing this will make sure
it can run the ANOVA correctly.

Convert your IV to a factor and your DV to a numeric in the code chunk
below using the `mutate()` function.

```{r convert IV and DV}



```

## Explore our data (summary and visualisation)

First, calculate some descriptive statistics to understand the data
better.

In the code chunk below, create a variable called `sum_dat`.

This `sum_data` will take `dat2` and then (remember the pipe operator?):

-   Use `GroupBy()` to group scores by our independent variable.
-   Use `summarise()` to calculate `mean`, `sd`, and `se` scores.

```{r create sum_data}


```

Once you have down that, visualize the data by uncommenting the code
below (remove all \`#\` and run it.

```{r visualise differences}

#ggplot(sum_dat, aes(x = Condition, y = mean, fill = Condition))+
#  stat_summary(fun = "mean", geom = "bar", show.legend = FALSE)+
#  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.25)+
#  scale_y_continuous(limits = c(0,7), 
#                     breaks = c(0,1,2,3,4,5,6,7), 
#                     name = "Intrusive-Memory Frequency (Mean for the Week")+
#  scale_x_discrete(labels = c("No-task control", "Reactivation plus Tetris", "Tetris #only",
#                                "Reactivation only"))

```

Looking at your descriptive statistics and visualization, what does your
data tell you so far about the research question?

## One-Way Anova

Run the one-way ANOVA using the `aov_ez` function. Save it to the
variable `ANOVA1`.

`aov_ez()` will likely produce some messages that look like errors, do
not worry about these, they are just letting you know what it's done.

```{r one-way aov, echo=FALSE}



```

We can use the `tidy()` function to make the output easier to work with.
Uncomment the code below and run it to transform the output. Don't worry
about the warning message, it is just telling you it doesn't know how to
automatically rename the columns so it will keep the original names.

```{r}

#ANOVA1 <- (ANOVA1$anova_table) %>% tidy()

```

Based on the output, answer the followning questions in this markdown
file (just create a new line under each question).

-   Is there the overall effect of Condition significant?

-   What is the F-statistics to 2 decimal places?

-   According to the rules of thumb, is the effect size is small,
    medium, or large?

## Check Parametric Assumptions

**Normality**

In the code chunk below, run a \`qqPlot\` and Shapiro-Wilk test to
assess if our residuals are normally distributed.

```{r normality, echo=TRUE}

```

*Is our data normally distributed?*

**Homogeneity of Variance**

In the code chunk below, run a Levene's test to assess whether there is
homogeneity of variance between the groups.

```{r homogeneity}

```

*Does the data display homogeneity of variance?*

## Running post-hoc tests

We are going to use the same approach as the lecture and the example, by
using the `emeans()` function.

In the code chunk below, use the `emmeans()` to run the comparisons.
Save the comparisons to the variable name `AOV1_pairwise.`

Then use the `tidy()`function to make it `AOV1_pairwise` easier to work
with. Use \`mutate()\` to convert the p-value to 3 decimal places. Save
it to the variable name `ANOV1_contrasts`

```{r post-hoc}

```

### **Write-up**

Based on your results, write up a conclusion of the paper in relation to
the research question below.
