---
title: "Anova Workshop Example"
author: "Ryan Donovan"
date: "2024-01-31"
output:
  html_document: default
  word_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Research Topic

In recent years, there has been a growing interest in exploring the
therapeutic potential of psychoactive treatments on mental health. This
study focuses on the effects of psychoactive substances like Psilocybin
and DMT on personality traits and emotional well-being, with a
particular emphasis on Neuroticism, which is associated with various
psychological disorders. We aimed to investigate whether the
administration of Psilocybin and DMT had observable impacts on
participants' Neuroticism levels. To do so, we recruited 60 participants
and randomly assigned them to three groups: Psilocybin, DMT, or Placebo.

Our goal was to determine if these substances influenced Neuroticism,
potentially offering insights into their therapeutic potential for
addressing emotional well-being and related mental health concerns.

**Research Question**: Is there an effect of psychoactive treatment on
Neuroticism?

***Our task***: Run a One-Way ANOVA to see if there is an effect of
Psychoactive treatment on Neuroticism.

## Steps to running an One-Way ANOVA

To conduct one-way ANOVA you should follow this general procedure:

1.  Prepare your analysis: Call in your data and any packages you need.
    Identify your IV and DV and convert them as appropriate.

2.  Explore your data: compute descriptive statistics and visualisations
    to assess whether there seems to be any difference between groups.

3.  Run the One-Way Anova.

4.  Check to see whether your data meets parametric assumptions.

5.  Compute contrasts or post hoc tests: having conducted the main
    ANOVA, you can follow it up with either contrasts or post hoc tests.
    For these sessions, it will be post hoc.

We will work through these steps in turn.

## Prepare your analysis

### Call Packages we need

```{r set-up, message = FALSE}

#remember to use install.packages("package name here") for any packages that do not load. 

library(patchwork)
library(car)
library(broom)
library(afex)
library(tidyverse)
library(pwr) 
library(lsr)
library(emmeans)
library(performance)

```

### Call in our Data

Download the **`psycho.csv`** dataset from Canvas, and then call it into
R, assigning it to the variable **`mydat`**. Use the **`head()`** and
**`summary()`** functions to inspect the data.

```{r}

mydat <- read.csv("psycho.csv")

head(mydat) #this will print out data for the first five participants. The head() function is really useful for quickly checking whether you have loaded in the data correctly. 

summary(mydat) #this will compute some quick descriptive statistics for us

```

There is not much information we are actually trying to ascertain from
`head()` or `summary()`, they give us a quick birds eye view of the
data - this can be really helpful to make sure you have loaded the
correct data set!

### Identify your IV and DV

Before we explore our data, we need to tell `R` which what are our
independent variable and dependent variables. Doing this will make sure
it can run the ANOVA correctly.

In this study, **`Treatment`** is the IV, and **`Neuroticism`** is the
DV. We need to convert **`Treatment`** to a factor and **`Neuroticism`**
to numeric data.

We can do this using the `mutate()` function. Remember that mutate()
takes an existing column within a dataset and performs an operation on
it. In this case, we are taking the `Treatment` and `Neuroticism`
columns in `mydat`, and converting them to a `factor` and `numeric` data
type.

```{r}
mydat <- mydat %>%
  mutate(Treatment = as.factor(Treatment)) %>%   #For the ANOVA to run properly our grouping (IV) needs to be set as a factor...
  mutate(Neuroticism = as.numeric(Neuroticism))  #... and our DV needs to be numeric


str(mydat) #we can check whether it was successful
```

Now we are ready to explore our data.

## Explore your data

First, calculate some descriptive statistics to understand the data
better.

```{r}

summary_stats <-mydat%>% #create a new variable called summary_stats. To create this variable, take the mydat dataset AND THEN (i.e., %>%)...
  group_by(Treatment)%>% #group scores based on whether they receive dmt, psilocybin, or a placebo AND THEN...
  summarise(mean = mean(Neuroticism), #calculate the mean score per group on Neuroticism
            sd = sd(Neuroticism), #calculate the sd score per group on Neuroticism 
            se = sd/sqrt(length(Neuroticism))) #calculate the standard error score per group on Neuroticism


summary_stats #print out the new variable

```

We can also visualise the differences

```{r Visualisation}
ggplot(mydat, aes(x = Treatment, y = Neuroticism))+
  geom_violin(trim = FALSE)+
  geom_boxplot(width = .2)
```

Okay, so we can see that `Psilocybin` scores higher than `Placebo`,
which scores higher than `DMT`. Maybe our participants had a
particularly hard trip.

Let's check whether to see if these differences are significantly
different.

## One-Way Anova

The following code will compute our one way ANOVA.

```{r one-way aov, echo=FALSE}
ANOVA1 <- aov_ez(id = "Participant_ID", 
              dv = "Neuroticism", 
              between = "Treatment", #what our IV is
              es = "pes", 
              type = 3, 
              include_aov = TRUE,
              data = mydat) 


anova(ANOVA1) 

ANOVA1_output <- (ANOVA1$anova_table) %>% tidy() 
```

### Code Explanation

Let's unpack what is going on here code wise:

1.  **`ANOVA1 <- aov_ez(...)`**: This line of code is creating a new R
    object named **`ANOVA1`**. This object will store the results of the
    one-way ANOVA analysis.

2.  **`id = "Participant_ID"`**: This part specifies the column in your
    dataset that contains the participant IDs. It helps identify which
    data points belong to each participant.

3.  **`dv = "Neuroticism"`**: Here, you specify the dependent variable

    (DV) of your analysis, which is `Neuroticism` in this case. This is
         the variable you want to analyze to see if it differs
         significantly between the treatment groups.

4.  **`between = "Treatment"`**: This indicates the independent variable

    (IV) you are testing, which is "Treatment" in your dataset. The
         one-way ANOVA will assess whether there are significant
         differences in Neuroticism scores between the different
         treatment groups.

5.  **`es = "pes"`**: This argument specifies that you want to compute
    the effect size in terms of partial eta-squared (pes). Effect size
    measures the strength of the relationship between variables, and
    eta-squared is commonly used in ANOVA to quantify this effect.

6.  **`type = 3`**: The **`type`** argument specifies the type of sum of
    squares used in the ANOVA. Type 3 is often used for balanced
    designs.

7.  **`include_aov = TRUE`**: This indicates that you want to include
    the standard ANOVA table as part of the output. It provides
    additional information about the analysis.

8.  **`data = mydat`**: Finally, you specify the dataset (**`mydat`**)
    on which you want to perform the ANOVA analysis.

The next line, **`ANOVA1`**, simply prints out the results of the ANOVA
analysis stored in the **`ANOVA1`** object. It will display information
about the main ANOVA, including the F-statistic, degrees of freedom, and
p-value.

Following that, **`anova(ANOVA1)`** performs a more detailed ANOVA
analysis and displays the ANOVA table, which includes information about
the sum of squares, mean squares, F-statistic, and p-value. This table
helps you determine if there are significant differences in
`Neuroticism` scores among the treatment groups.

Finally, **ANOVA1_output <- (ANOVA1$anova_table) %>% tidy()**

Just cleans it up so it's easier to work with. 


### Data Explanation

-   The main effect of "Treatment" on Neuroticism scores was
    statistically significant (F(2, 57) = 9.05, p \< 0.001). This
    indicates that there are significant differences in Neuroticism
    scores among the treatment groups.

-   The effect size (partial eta-squared, ges) is 0.241, which suggests
    that approximately 24.1% of the variance in Neuroticism scores can
    be attributed to the differences in treatment groups.

## Check Parametric Assumptions

```{r One-Way ANOVA assumptions, echo=TRUE}
qqPlot(ANOVA1$aov$residuals)
shapiro.test(ANOVA1$aov$residuals)
test_levene(ANOVA1)
```

Okay, so we can see that our data has met parametric assumptions and
there is a significant effect of psychoactive treatment on
`Neuroticism`. But where does it lie?

## Running post-hoc tests

```{r One-way ANOVA post-hoc}
AOV1_pairwise <-emmeans(ANOVA1, pairwise ~ Treatment, adjust = "bonferroni")

```

1.  **`AOV1_pairwise <- emmeans(...)`**: This part of the code enables
    us to calculate and stores pairwise comparisons of group means using
    the **`emmeans`** function.

2.  **`ANOVA1`** is the previously defined object containing the results
    of the one-way ANOVA.

3.  **`pairwise ~ Treatment`** specifies that you want to compare the
    treatment groups

4.  **`adjust = "bonferroni"`** indicates that you want to apply the
    Bonferroni correction to control for familywise error rate in
    multiple comparisons.

This step is necessary because after finding a significant result in the
ANOVA, you often want to determine which specific groups differ
significantly from each other. The Bonferroni correction adjusts the
p-values to account for the increased risk of Type I errors when
conducting multiple comparisons.

```{r}

AOV1_contrasts <- AOV1_pairwise$contrasts %>% tidy() %>% # tidy the output
  mutate(adj.p.value = round(adj.p.value, digits = 3)) # changes the p value to be in 3 digits (aka what we are used to)
AOV1_contrasts

```

In these pairwise comparisons between treatment groups for Neuroticism
scores:

-   The DMT group did not show a statistically significant difference
    compared to the Placebo group (p = 1.000), indicating similar
    Neuroticism scores.

-   However, the DMT group exhibited a statistically significant
    difference compared to the Psilocybin group (p ≤ 0.001), suggesting
    that their Neuroticism scores differed significantly.

-   Likewise, the Placebo group showed a statistically significant
    difference compared to the Psilocybin group (p = 0.005), indicating
    that their Neuroticism scores were significantly different.

## Write Up

A One-Way ANOVA test was conducted to investigate if there was a
significant effect of psychoactive treatment on Neuroticism This test
was selected as the data was scale data, there were equal numbers
between groups, the residuals were normally distributed, and there was
homogeneity of variance. The results of the One-Way ANOVA demonstrated
that there was a significant effect of psychoactive treatment on
Neuroticism scores, F(57) = 9.048, p ≤ .001, ƞ² = .24. Post-hoc
comparisons revealed that participants administered with Psilocybin
experienced a significant increase in Neuroticism than participants
administered with DMT (p ≤ .001) or a placebo (p = .005).
