---
title: "ANOVA"
author: "CiaraEgan"
date: "2023-01-27"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Call Packages we need

```{r set up}
library(patchwork)
library(car)
library(broom)
library(afex)
library(tidyverse)
library(pwr) 
library(lsr)
library(emmeans)
library(performance)


```

## Call in data

Todays examples will come from the TMDB movie database 

```{r data, echo=TRUE}
mydat <- read_csv("clinicaltrials.csv") #Reads the csv file into R and saves it as an object called "mydat"
head(mydat)
summary(mydat)
```
```{r data format}
mydat <- mydat %>%
  mutate(drug = as.factor(drug)) %>%   #For the ANOVA to run properly our grouping (IV) needs to be set as a factor...
  mutate(mood.gain = as.numeric(mood.gain))  #... and our DV needs to be numeric

```

```{r descriptive statistics}
summary_stats <-mydat%>%
  group_by(drug)%>%
  summarise(mean = mean(mood.gain),
            sd = sd(mood.gain),
            se = sd/sqrt(length(mood.gain)))
summary_stats
```

## One-way ANOVA
Now we are going to run a One-way ANOVA to see if there is a difference in mood gain between the two drugs & placebo
```{r one-way aov, echo=FALSE}
ANOVA1 <- aov_ez(id = "ID", # the column containing the participants IDs
              dv = "mood.gain", # the DV 
              between = "drug", # the between-subject variable
              es = "pes", # sets effect size to partial eta-squared
              type = 3, # this affects how the sum of squares is calculated, set this to 3
              include_aov = TRUE,
              data = mydat) # We are creating an object called "ANOVA1" which is the results of a one-way ANOVA looking at the difference in revenue between film genres.
anova(ANOVA1)
ANOVA1_output <- (ANOVA1$anova_table) %>% tidy() 
```

## Checking our assumptions
1. The DV is interval or ratio data
2. The observations should be independent
3. The residuals should be normally distributed
4. There should be homogeneity of variance between the groups

```{r One-Way ANOVA assumptions, echo=TRUE}
qqPlot(ANOVA1$aov$residuals)
shapiro.test(ANOVA1$aov$residuals)
test_levene(ANOVA1)
```


## Running post-hoc tests
So now we want to see where the differences lie?

```{r One-way ANOVA post-hoc}
AOV1_pairwise <-emmeans(ANOVA1, pairwise ~ drug, adjust = "bonferroni")
AOV1_contrasts <- AOV1_pairwise$contrasts %>% tidy() %>% # tidy the output
  mutate(adj.p.value = round(adj.p.value, digits = 3)) # changes the p value to be in 3 digits (aka what we are used to)
AOV1_contrasts
```
```{r Visualisation}
ggplot(mydat, aes(x = drug, y = mood.gain))+
  geom_violin(trim = FALSE)+
  geom_boxplot(width = .2)
```



## Factorial ANOVA 
Scenario:
Now we want to compare the efficacy of both drugs and therapy (no therapy vs CBT) on participants mood, and compare across the three drug types also.

```{r data set up for factorial ANOVA, echo=TRUE}
mydat <- mydat %>%
  mutate(therapy = as.factor(therapy)) 
summary(mydat)
```
```{r factorial ANOVA summary stats}
summary_stats2 <-mydat%>%
  group_by(drug, therapy)%>%
  summarise(mean = mean(mood.gain),
            sd = sd(mood.gain),
            se = sd/sqrt(length(mood.gain)))
summary_stats2
```


## Factorial ANOVA
Now we are going to run a Factorial ANOVA. 
```{r factorial aov, echo=FALSE}

ANOVA2 <- aov_ez(id = "ID", # the column containing the film IDs
              dv = "mood.gain", # the DV 
              between = c("drug", "therapy"), # the between-subject variables
              es = "pes", # sets effect size to partial eta-squared
              type = 3, # this affects how the sum of squares is calculated, set this to 3
              include_aov = TRUE,
              data = mydat) # We are creating an object called "ANOVA1" which is the results of a one-way ANOVA looking at the difference in revenue between film genres.
anova(ANOVA2)
ANOVA2_output <- (ANOVA2$anova_table) %>% tidy()
```


## Factorial ANOVA Assumptions
Now we are going to check our assumptions (Same as above).

1.The DV is interval or ratio data
2.The observations should be independent
3.The residuals should be normally distributed
4.There should be homogeneity of variance between the groups

```{r Factorial ANOVA assumptions, echo=TRUE}
qqPlot(ANOVA2$aov$residuals)
shapiro.test(ANOVA2$aov$residuals)
test_levene(ANOVA2)
```


```{r Violin plots}
ggplot(mydat, 
       aes(x = drug , y = mood.gain, fill = therapy))+
  geom_violin(trim = FALSE, 
              alpha = .4)+
  geom_boxplot(position = position_dodge(.9), 
               width = .2, 
               alpha = .6)+
  scale_fill_viridis_d(option = "E")+
  stat_summary(fun = "mean", geom = "point",
               position = position_dodge(width = 0.9)) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = .1,
               position = position_dodge(width = 0.9)) +
  theme_minimal()
```





```{r interaction plots}
ggplot(summary_stats2, aes(x = therapy, y = mean, group = drug, shape = drug)) +
  geom_point(size = 3) +
  geom_line(aes(linetype = drug))+
  theme_classic()
```

